Approximation as Differentiation: Notes

### 2025-06-03 notes/language.tex
Reference language sections.
Restrict to closed function types for now.

We introduce a standard total functional language with a (reasonably) expressive type
system~\cite{crole94,pitts01,santocanale02}, over a set $\PrimTy$ of primitive types $\rho$ and sets
$\PrimOp^\rho_{\rho_1,\ldots,\rho_n}$ of primitive operations.

\subsubsection{Syntax}

\input{fig/positivity}
\input{fig/syntax}
\input{fig/typing}

@@

\subsubsection{Semantics}

@@

\begin{definition}[$\mu$-polynomial]
Suppose $\cat{C}$ a category with finite coproducts and finite products. Define $\muPoly_{\cat{C}}$ to be the
smallest subcategory of $\Cat$ generated inductively by the rules in \figref{mu-polynomial}.
\end{definition}

@@

For the sake of a ``default'' (baseline) semantics, fix a bicartesian closed category $\cat{C}$ with finite
products $(\times, \One)$, finite coproducts $(\coprod, \Zero)$, exponentials $\internalHom{X}{Y}$, evaluation
morphisms $\eval_{X,Y}$ and currying isomorphisms $\lambda_{X,Y,Z}$, plus the following additional structure:
\begin{enumerate}
\item a strong monad $\mathcal{L}$ (lifting) with unit $\eta_X: X \to \mathcal{L}(X)$, multiplication $\mu_X:
\mathcal{L}^2(X) \to \mathcal{L}(X)$ and strength $\mathsf{st}_{X,Y}: X \times \mathcal{L}(Y) \to
\mathcal{L}(X \times Y)$;
\item $\mu$-polynomial endofunctors given by $\muPoly_{\cat{C}}$;
\item for each primitive type $\rho \in \PrimTy$, an object $\sem{\rho}$ of $\cat{C}$;
\item for each primitive operation $\phi \in \PrimOp^\rho_{\rho_1,\ldots,\rho_n}$, a morphism $\sem{\rho}:
\sem{\rho_1} \times \ldots \times \sem{\rho_n} \to \sem{\rho}$.
\end{enumerate}
\figref{default-semantics:types} gives the default interpretation of types $\Delta \vdash \tau$ with
$\length{\Delta} = n$ as multifunctors $\cat{C}^n \to \cat{C}$ (or as objects of $\cat{C}$ in the case $n =
0$). \figref{default-semantics:terms} gives the default interpretation of terms $\Gamma \vdash t: \tau$ as
morphisms $\sem{\Gamma} \to \sem{\tau}$.

>>>>

We introduce a standard total functional language with a (reasonably) expressive type
system~\cite{crole94,pitts01,santocanale02}, over a set $\PrimTy$ of primitive types $\rho$ and sets
$\PrimOp^\rho_{\rho_1,\ldots,\rho_n}$ of primitive operations $\phi$.

\subsubsection{Syntax}
\label{sec:language:syntax}

% \input{fig/positivity} -- follow nunes2023 instead
\input{fig/syntax}
\input{fig/typing}

@@

\subsubsection{Semantics}
\label{sec:language:semantics}

@@

\begin{definition}[$\mu$-polynomial]
Suppose $\cat{C}$ a category with finite coproducts $(\coprod, \Zero)$ and finite products $(\times, \One)$.
Define $\muPoly_{\cat{C}}$ to be the smallest subcategory of $\Cat$ generated inductively by the rules in
\figref{mu-polynomial}.
\end{definition}

@@

For the sake of a ``default'' (baseline) semantics, fix a bicartesian closed category $\cat{C}$ with finite
products $(\times, \One)$, finite coproducts $(\coprod, \Zero)$, exponentials $\internalHom{X}{Y}$, evaluation
morphisms $\eval_{X,Y}$ and currying isomorphisms $\lambda_{X,Y,Z}$, plus the following additional structure:
\begin{enumerate}
\item a strong monad $\mathcal{L}$ (lifting) with unit $\eta_X: X \to \mathcal{L}(X)$, multiplication $\mu_X:
\mathcal{L}^2(X) \to \mathcal{L}(X)$ and strength $\mathsf{st}_{X,Y}: X \times \mathcal{L}(Y) \to
\mathcal{L}(X \times Y)$;
\item $\mu$-polynomial endofunctors given by $\muPoly_{\cat{C}}$;
\item for each primitive type $\rho \in \PrimTy$, an object $\sem{\rho}$ of $\cat{C}$;
\item for each primitive operation $\phi \in \PrimOp^\rho_{\rho_1,\ldots,\rho_n}$, a morphism $\sem{\rho}:
\sem{\rho_1} \times \ldots \times \sem{\rho_n} \to \sem{\rho}$.
\end{enumerate}

\figref{default-semantics:types} gives the default interpretation of types $\Delta \vdash \tau$ with
$\length{\Delta} = n$ as multifunctors $\cat{C}^n \to \cat{C}$ (or as objects of $\cat{C}$ in the case $n =
0$). \figref{default-semantics:terms} gives the default interpretation of terms $\Gamma \vdash t: \tau$ as
morphisms $\sem{\Gamma} \to \sem{\tau}$.
